---
title: "Activity Data Analysis"
author: "Nitish Dobhal"
date: "July 21, 2017"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=FALSE)
```

##Introduction##
This data is a part of the research in the "Quantified Self" movement, in which the individuals through use of modern devices can track their personal activities as well as health. This data is then processed and analyzed to better the health of the individual.

The given data was collected for a duration of 2 months and recorded activity i.e steps taken every 5 minute. This work aims to get the basic statistics in form of mean and medians of data, as well a comparison of activities during weekdays and weekends.

##Loading Data##
The data is loaded from the [link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) provided within the assignment page. The following code makes sure that the data files are not downloaded unless they are not found in the working directory.

```{r Loading Data}
fname<-"ActivityData.zip"
if(!file.exists(fname)) {
  url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(url, destfile = "ActivityData.zip")
}
unzip("ActivityData.zip")
a_data<-read.csv("activity.csv")
```

##Transforming Data##
Within the 60 day monitoring period, lot of sensor data is missing. The first step is to examine the results without imputing the missing data. Following code 
```{r RemoveNA}
a_data<-a_data[!is.na(a_data$steps),]
a_data<-droplevels(a_data)
```

##Calculating Basic Statistics (without Imputation)

###a)Total Steps each day###
For calculating the total number of steps, we will apply the tapply function on the 'steps' vector with factor variable 'date' and use sum function.
```{r TotalSteps}
t_steps<-tapply(a_data$steps,a_data$date,sum,simplify = T)
summary(t_steps)
```

###b)Histogram of Total Steps per day###
```{r HistWNA, echo=T}
mean_total<-mean(t_steps)
names(mean_total)<-"Mean"
med_total<-median(t_steps)
names(med_total)<-"Median"
hist(t_steps,breaks=20,xlab = "Total Steps Each Day",main="Histogram of Total Steps per day for 2 months",col="lightblue")
abline(v=mean_total,lty=3,lwd=3,col="red")
text(mean_total-500,6,"Mean",srt=90)
```


###c)Mean & Median of the total steps taken each day###
```{r , echo=T}
mean_total
med_total
```

###d)Average daily activity pattern###
'tapply' function is used to implement summation over the 'steps' vector with 'intervals' as factor. The output is an array which is coerced to a data frame. Proper names are given to the column of the result. A new vector is added to the result i.e 'interval' which is extracted from the row names of the output. The row names are then set to NULL to restore the default numbering.

```{r , echo=TRUE}
time_steps<-as.data.frame(tapply(a_data$steps,a_data$interval,mean,simplify = T))
names(time_steps)<-"steps"
time_steps$interval<-as.numeric(row.names(time_steps))
row.names(time_steps)<-NULL
plot(y=time_steps$steps,x=time_steps$interval,type="l",xlab = "Time (minutes)",ylab="Average Steps",main = "Time Variation of Steps taken during a day",col="red")

```

###e)Finding 5-minute interval with most steps on an average###
```{r, echo=TRUE}
max_step_i<-names(which(time_steps$steps==max(time_steps$steps)))
max_step_i
```

##Calculating Basic Statistics (with Imputation)
A simple imputation using the **mean** of steps taken for a particular interval is done. All the missing rows are first extracted into a separate data frame. 'match' function is used to lookup the 'interval' for a particular row in the 'time_steps' variable (used earlier to store average steps for each interval) and corrosponding value of 'steps' is returned. This is repeated over all the rows using 'sapply' function. The rows with imputed data is then merged back with the original data

```{r}
a_data2<-read.csv("activity.csv")
missing_V<-a_data2[is.na(a_data2$steps),]
pctage_m<-mean(is.na(a_data2$steps))
missing_V$steps<-sapply(missing_V$interval,function(x)time_steps$steps[match(x,time_steps$interval)])
a_data2[is.na(a_data2$steps),]<-missing_V
```

The original percentage of missing values is shown below. The new imputed data is stored in 'a_data2'.
```{r,echo=T}
paste0(as.character(round(pctage_m,4)*100),"%")
```
###a)Histogram of Total Steps per day with imputation###

```{r HISTWONA, echo=T}
t_steps2<-tapply(a_data2$steps,a_data2$date,sum,simplify = T)
mean_total2<-mean(t_steps2)
names(mean_total2)<-"Mean"
med_total2<-median(t_steps2)
names(med_total2)<-"Median"

hist(t_steps2,breaks=20,xlab = "Total Steps Each Day",main="Histogram of Total Steps per day for 2 months",col="lightblue")
abline(v=mean_total2,lty=3,lwd=3,col="red")
text(mean_total2-500,6,"Mean",srt=90)
```

```{r, echo=F}
paste0("Original Mean without Imputation: ",as.character(round(mean_total,2)))
paste0("New Mean with Imputation: ",as.character(round(mean_total2,2)))
paste0("Original Median without Imputation: ",as.character(med_total))
paste0("New Median with Imputation: ",as.character(round(med_total2,2)))
```

There is no difference in the new and old mean because the imputed values are derived from the original mean value.  The median has increased as there are more values weighted towards the mean value than before.

##Difference in activity between Weekdays & Weekends##
To find out if a given date is a weekday or weekend, 'weekdays()' function has been used. Since this function accepts an argument of 'Date' type, the 'date' vector of 'a_data2' is first coerced into a date object. Consequently a new logical vector 'weekd' is added indicating if a given day is weekend(True) or not(False). 'tapply' function is then applied on both groups separately to extract the time series data into two separate variables- 'time_steps_we' & 'time_steps_wd'. 
```{r, fig.width=9,fig.height=9}
a_data2$date<-as.Date(a_data2$date,"%Y-%m-%d")
a_data2$weekd<-grepl("S.+",weekdays(a_data2$date))
time_steps_we<-as.data.frame(tapply(a_data2[a_data2$weekd,1],a_data2[a_data2$weekd,3],mean,simplify = T))
time_steps_wd<-as.data.frame(tapply(a_data2[!a_data2$weekd,1],a_data2[!a_data2$weekd,3],mean,simplify = T))
names(time_steps_we)<-"steps"
names(time_steps_wd)<-"steps"
time_steps_we$interval<-as.numeric(row.names(time_steps_we))
time_steps_wd$interval<-as.numeric(row.names(time_steps_wd)) 
par(mfrow=c(2,1),mar=c(4,2,2,1))
plot(time_steps_wd$steps,x=time_steps_wd$interval,type="l",xlab= "Time (minutes)",ylab="Average Steps",main = "Time Variation of Steps taken during Weekdays",col="red")
plot(time_steps_we$steps,x=time_steps_we$interval,type="l",xlab= "Time (minutes)",ylab="Average Steps",main = "Time Variation of Steps taken during Weekend",col="blue")
```


